Aspects of Dealing with Imperfect Data in Temporal Databases

### 2012-03-16 trunk/book/sources/temporalDatabases.tex
Still cool...

%
% Temporal databases
%
A temporal database can generally be seen as a database that manages some temporal aspects in its schema \cite{etzion1998},~\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. In subsection \ref{subsec:td-general-concepts}, some main concepts and properties concerning temporal databases and their definitions are presented and explained. In subsections \ref{subsubsec:primary-key} and \ref{subsubsec:consistency}, some main issues of relational temporal databases are presented and discussed. Finally, subsection \ref{Comm-temp} presents an overview of some commercial temporal database systems.

@@

Thus, a temporal database will contain \emph{temporal values}, i.e. values representing (indications of) time. Temporal values in a temporal database can be classified into the following types based on their interpretation and modelling purpose. The definitions and explanations of these types can be found in \cite{Dyreson1994} and \cite{Nascimento95decisiontime} and more information can be found in \cite{Jensen:1991:IIM:627283.627484}, \cite{Snodgrass:1984:TQL:588011.588041} and \cite{Nascimento95decisiontime}.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Valid Time}\\
The \emph{\textbf{valid time}} (VT) of a fact is the time when the fact is true in the modeled reality.
\end{definition}

\begin{definition}\textbf{Transaction Time}\\
A database fact is stored in a database at some point in time, and after it is stored, it is current until logically deleted. The \emph{\textbf{transaction time}} (TT) of a database fact is the time when the fact is current in the database and may be retrieved.
\end{definition}

\begin{definition}\textbf{Decision Time}\\
\emph{\textbf{Decision time}} (DT) denotes the time when an event was decided to happen.
\end{definition}

\begin{definition}\textbf{User-defined Time}\\
\emph{\textbf{User-defined time}} (UDT) is an uninterpreted attribute domain of date and time.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

\emph{Valid times} are usually provided by the user, whereas \emph{transaction times} are usually system-generated and -supplied \cite{Dyreson1994}. Temporal values of type UDT are not given any extraordinary interpretation and have thus no extraordinary query language support \cite{Dyreson1994}.

A \emph{temporal database} can now formally be defined as follows \cite{Dyreson1994}.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Temporal Database}\\
A \emph{\textbf{temporal database}} supports some aspect of time, not counting user-defined time.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

In a relational temporal database, temporal values will of course be in the tuples of the extentions of temporal relations:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Valid-time Relation}\\
A \emph{\textbf{valid-time relation}} is a relation with exactly one system supported valid time \cite{Dyreson1994}.
\end{definition}

\begin{definition}\textbf{Transaction-time Relation}\\
A \emph{\textbf{transaction-time relation}} is a relation with exactly one system supported transaction time \cite{Dyreson1994}.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

A \emph{valid-time}, respectively \emph{transaction-time} \emph{relational database} is now defined as containing one or more valid-time, respectively transaction-time relations \cite{Dyreson1994}. Next to this, \emph{bitemporal} relational databases contain both valid time and transaction time~\cite{Dyreson1994} and tritemporal databases contain valid time, transaction time and decision time~\cite{Nascimento95decisiontime}.

%Database models can also be classified into \emph{bi-temporal} (both valid and transaction-time) or \emph{tri-temporal}  (bi-temporal and decision time) models.

@@

\subsection{\label{subsubsec:primary-key}Primary Keys in Valid-time Relation Design}
Generally, when designing a relation based on a relational database model, a subset of the relation's attribute set is ususally chosen as primary key. The values of a tuple for these attributes will then uniquely determine this tuple, hence no two distinct tuples may have the exact same values for every attribute in this primary key. Next to attributes unrelated to time, a valid-time relation schema will typically contain one or more attributes which model the valid-time aspects and behavior of the real objects and concepts modelled by the relation schema. In this work, these attributes are called \emph{valid-time attributes}. In valid-time relation extentions, distinct tuples can exist containing the exact same values for every attribute except the valid-time attributes. These distinct tuples represent distinct versions of the same real object or concept, valid during different time periods. To allow the existence of such tuples when designing a valid-time relation using a relational database model, the most common solution is to include the valid-time attributes in the primary key.
%One of the first tasks in the design of a database is the choice for the primary key. In a temporal database, the most common solution for the choice of the primary key is to extend the primary key with the temporal value. Nevertheless, when dealing with uncertainty in the temporal values, the solution is usually to create a version identifier otherwise the primary key have uncertainty which should be avoided.

@@

\begin{example}
\label{ex:pk}
Consider the example valid-time relation visualized in table \ref{table:example-database}, which models when certain people worked as employees in a certain company and under whose supervision they worked during that time. The valid-time attributes `Start' and `End' describe the year when an employee started, respectively finished working for the company. For example, the last tuple visualized in table \ref{table:example-database} represents that the employee represented by this tuple started working for the company in 2005 and finished in 2009. The attributes `Name', `Birthday' and `Supervisor' describe respectively the name, birthday date and unique identifier of the supervisor of an employee in the time during which he or she worked for the company. When correct, the date of an employee's birthday never changes and as such, the modelling of birthday dates has no effect on the database consistency. The `Birthday' attribute thus describes UDT values. The attribute `ID' describes employee identifiers. For each tuple, this identifier (a number) uniquely identifies the employee represented by the tuple.

Now consider $\{$ID$\}$ being the primary key and consider the company wanting to hire Sarah again in 2010. This would be represented by another tuple in the relation, containing value 4 for attribute `ID'. The existence of such a tuple is of course not allowed by the primary key, because it would mean the existence of two distinct tuples containing value 4 for attribute `ID'. This problem can now be solved by defining a new primary key: $\{$ID, Start, End$\}$, which allows for the existence of distinct tuples with value 4 for attribute `ID', as long as they have different values for attributes `Start' or `End'. The resulting relation is shown in table~\ref{table:example-database-with-new-pk}.
%Consider the example database visualized in table \ref{table:example-database}. It contains data representing employees in a company, and two temporal values which represent a valid-time interval. Consider now that we want to hire Sarah again. That is not possible because of the primary key (ID) does not allow to insert again the row Sarah but with a different start and end time. In some models this is resolved by adding to the primary key both values Start and End years. The resulting database allows to insert a new row where Sarah the year 2010. But this modification also allows to insert spurious values e.g. inconsistent time periods (see last row in table \ref{table:example-database-with-new-pk}).

@@

\begin{example}
\label{ex:prob}
Consider the example valid-time relation visualized in table \ref{table:erconsistency}, which is based on the relation visualized in table~\ref{table:example-database}. The primary key is again $\{$ID, Start, End$\}$. The last record in the relation represents a person named `Sarah' started working for the company in 2007 and finished in 2008, with supervisor `John'. However, the fourth record represents the same person (the value for attribute `ID' is the same) started working for the company in 2005 and finished in 2009, with supervisor `Maria'. The intention is clear: Sarah worked in the company from 2005 to 2009, first for Maria, then for John, then again for Maria. It is of course possible for an employee to change supervisors, but it is of course impossible for a person to start working in the same company twice at different times, for different supervisors, without stopping to work for one in between, as it is impossible to stop working for a supervisor twice at different times, without working for another one in between. The valid-time information represented by the last record is clearly not consistent with the valid-time information represented by the fourth record, or vice versa.
\end{example}

@@

%The consistence mechanism in a temporal database is usually re-defined. The main problem is the insertion of spurious values in the database, as shown in example \ref{ex:pk}, in table \ref{table:example-database-with-new-pk}. The most usual solution is to re-define the DML (\emph{Data Manipulation Language}). e.g. the \emph{update} sentence is redefined as two sentences an update and a create sentence as illustrated in the following example.

@@

\subsection{\label{Comm-temp}Commercial Temporal Database Systems}
Several commercial temporal DBMS exist. Table \ref{table:commercial-temporal-db} gives an overview of some of the more well-known temporal DBMS and provides references for more information.

Oracle workspace manager~\cite{oracle2009} and TimeDB~\cite{timedb2005} are libraries for dealing with time in OracleDB. On another note, TimeDB and Postgree Temporal~\cite{posgree2009} are similar: both are simple implementations that implement a subset of the Allen operators and some operations for the creation and manipulation of temporal attributes (valid time, transaction time or both times are supported). Teradata~\cite{teradata2011} is mainly a business intelligence system designed for data mining. Secondo~\cite{Dieker2000} is an extensible database system in which the core of the database may be replaced by a customized algebra. It is designed for non-standard applications and it supports both valid and transaction times.

@@

%\subsubsection{Oracle Workspace Manager}
%
%\begin{figure}
%\centering
%\includegraphics[scale=0.5]{graphs/workspaceTopology.eps}
%\caption{A sample topology in the workspace manager.}
%\label{fig:workspace-topology}
%\end{figure}
%
%Oracle workspace manager \cite{OraE118602} package allows to get several versions of the data in the same database. It is also possible to version only a table. The main benefits of this are two:
%
%\begin{itemize}
%\item
%Organization and optimization of data in a hierarchical frames: It is possible to create a frame of time (a workspace) and work only with that partial version of the data. It is provided a mechanism to merge data among different workspaces and to solve conflicts. The organization of data in these workspaces for very large tables result in smaller time access.
%\item
%Valid time as well as transaction time are managed by the system. Each update sentence makes a change in the versioning of a row.
%\end{itemize}
%
%When a table is versioned, the system creates a few tables and views as well as auxiliary structures to allow keep several version of the data, while keeping the primary key defined by the user.
%
%A workspace is a logical group of a set of changes (versions of tuples) and allows consistent access, thus the user always obtain the correct data version. Workspaces are ordered in a hierarchy. The top level is called LIVE workspace (see figure \ref{fig:workspace-topology}).  \\
%
%The system only makes a copy of a tuple if it is changed. In order to access different versions of the tuple, the context must be changed. Also, it is not necessarily to modify the SQL code to access to versioned tables.\\
%%When a table is versioned, it is renamed as table\_LT. Here is stored all the data for the information together metadata for the versioning. An auxiliary table is created with the workspace metadata with the name table\_AUX. A view is created with the name of the original table to allow querying without versioning.

>>>>

%
% Temporal databases
%
A temporal database can generally be seen as a database that manages some temporal aspects in its schema~\cite{etzion1998},~\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. In subsection~\ref{subsec:td-general-concepts}, some main concepts and properties concerning temporal databases and their definitions are presented and explained. In subsections~\ref{subsubsec:primary-key} and~\ref{subsubsec:consistency}, some main issues of relational temporal databases are presented and discussed. Finally, subsection~\ref{Comm-temp} presents an overview of some commercial temporal database systems.

@@

Thus, a temporal database will contain \emph{temporal values}, i.e. values representing (indications of) time. Temporal values in a temporal database can be classified into the following types based on their interpretation and modelling purpose. The definitions and explanations of these types can be found in~\cite{Dyreson1994} and~\cite{Nascimento95decisiontime} and more information can be found in~\cite{Jensen:1991:IIM:627283.627484},~\cite{Snodgrass:1984:TQL:588011.588041} and~\cite{Nascimento95decisiontime}.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Valid Time}~\cite{Dyreson1994}\\
The \emph{\textbf{valid-time}} (VT) of a fact is the time when the fact is true in the modeled reality.
\end{definition}

\begin{definition}\textbf{Transaction Time}~\cite{Dyreson1994}\\
A database fact is stored in a database at some point in time, and after it is stored, it is current until logically deleted. The \emph{\textbf{transaction-time}} (TT) of a database fact is the time when the fact is current in the database and may be retrieved.
\end{definition}

\begin{definition}\textbf{Decision Time}~\cite{Nascimento95decisiontime}\\
\emph{\textbf{Decision time}} (DT) denotes the time when an event was decided to happen.
\end{definition}

\begin{definition}\textbf{User-defined Time}~\cite{Dyreson1994}\\
\emph{\textbf{User-defined time}} (UDT) is an uninterpreted attribute domain of date and time.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

\emph{Valid times} are usually provided by the user, whereas \emph{transaction-times} are usually system-generated and -supplied~\cite{Dyreson1994}. Temporal values of type UDT are not given any extraordinary interpretation and have thus no extraordinary query language support~\cite{Dyreson1994}.

A \emph{temporal database} can now formally be defined as follows:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Temporal Database}~\cite{Dyreson1994}\\
A \emph{\textbf{temporal database}} supports some aspect of time, not counting user-defined time.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

In a relational temporal database, temporal values will of course be in the tuples of the extensions of temporal relations:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Valid-time Relation}~\cite{Dyreson1994}\\
A \emph{\textbf{valid-time relation}} is a relation with exactly one system supported valid-time.
\end{definition}

\begin{definition}\textbf{Transaction-time Relation}~\cite{Dyreson1994}\\
A \emph{\textbf{transaction-time relation}} is a relation with exactly one system supported transaction-time.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

A \emph{valid-time}, respectively \emph{transaction-time} \emph{relational database} is now defined as containing one or more valid-time, respectively transaction-time relations~\cite{Dyreson1994}. Next to this, \emph{bitemporal} relational databases contain both valid-time and transaction-time~\cite{Dyreson1994} and tritemporal databases contain valid-time, transaction-time and decision-time~\cite{Nascimento95decisiontime}.

%Database models can also be classified into \emph{bi-temporal} (both valid and transaction-time) or \emph{tri-temporal}  (bi-temporal and decision-time) models.

@@

\subsection{\label{subsubsec:primary-key}Primary Keys in Valid-time Relation Design}
Generally, when designing a relation based on a relational database model, a subset of the relation's attribute set is usually chosen as primary key. The values of a tuple for these attributes will then uniquely determine this tuple, hence no two distinct tuples may have the exact same values for every attribute in this primary key. Next to attributes unrelated to time, a valid-time relation schema will typically contain one or more attributes which model the valid-time aspects and behavior of the real objects and concepts modelled by the relation schema. In this work, these attributes are called \emph{valid-time attributes}. In valid-time relation extensions, distinct tuples can exist containing the exact same values for every attribute except the valid-time attributes. These distinct tuples represent distinct versions of the same real object or concept, valid during different time periods. To allow the existence of such tuples when designing a valid-time relation using a relational database model, the most common solution is to include the valid-time attributes in the primary key.
%One of the first tasks in the design of a database is the choice for the primary key. In a temporal database, the most common solution for the choice of the primary key is to extend the primary key with the temporal value. Nevertheless, when dealing with uncertainty in the temporal values, the solution is usually to create a version identifier otherwise the primary key have uncertainty which should be avoided.

@@

\begin{example}
\label{ex:pk}
Consider the example valid-time relation visualized in table~\ref{table:example-database}, which models when certain people worked as employees in a certain company and under whose supervision they worked during that time. The valid-time attributes `Start' and `End' describe the year when an employee started, respectively finished working for the company. For example, the last tuple visualized in table~\ref{table:example-database} represents that the employee represented by this tuple started working for the company in 2005 and finished in 2009. The attributes `Name', `Birthday' and `Supervisor' describe respectively the name, birthday date and unique identifier of the supervisor of an employee in the time during which he or she worked for the company. When correct, the date of an employee's birthday never changes and as such, the modelling of birthday dates has no effect on the database consistency. The `Birthday' attribute thus describes UDT values. The attribute `ID' describes employee identifiers. For each tuple, this identifier (a number) uniquely identifies the employee represented by the tuple.

Now consider $\{$ID$\}$ being the primary key and consider the company wanting to hire Sarah again in 2010. This would be represented by another tuple in the relation, containing value 4 for attribute `ID'. The existence of such a tuple is of course not allowed by the primary key, because it would mean the existence of two distinct tuples containing value 4 for attribute `ID'. This problem can now be solved by defining a new primary key: $\{$ID, Start, End$\}$, which allows for the existence of distinct tuples with value 4 for attribute `ID', as long as they have different values for attributes `Start' or `End'. The resulting relation is shown in table~\ref{table:example-database-with-new-pk}.
%Consider the example database visualized in table \ref{table:example-database}. It contains data representing employees in a company, and two temporal values which represent a valid-time interval. Consider now that we want to hire Sarah again. That is not possible because of the primary key (ID) does not allow to insert again the row Sarah but with a different start and end time. In some models this is resolved by adding to the primary key both values Start and End years. The resulting database allows to insert a new row where Sarah the year 2010. But this modification also allows to insert spurious values e.g., inconsistent time periods (see last row in table \ref{table:example-database-with-new-pk}).

@@

\begin{example}
\label{ex:prob}
Consider the example valid-time relation visualized in table~\ref{table:erconsistency}, which is based on the relation visualized in table~\ref{table:example-database}. The primary key is again $\{$ID, Start, End$\}$. The last record in the relation represents a person named `Sarah' started working for the company in 2007 and finished in 2008, with supervisor `John'. However, the fourth record represents the same person (the value for attribute `ID' is the same) started working for the company in 2005 and finished in 2009, with supervisor `Maria'. The intention is clear: Sarah worked in the company from 2005 to 2009, first for Maria, then for John, then again for Maria. It is of course possible for an employee to change supervisors, but it is of course impossible for a person to start working in the same company twice at different times, for different supervisors, without stopping to work for one in between, as it is impossible to stop working for a supervisor twice at different times, without working for another one in between. The valid-time information represented by the last record is clearly not consistent with the valid-time information represented by the fourth record, or vice versa.
\end{example}

@@

%The consistence mechanism in a temporal database is usually re-defined. The main problem is the insertion of spurious values in the database, as shown in example \ref{ex:pk}, in table \ref{table:example-database-with-new-pk}. The most usual solution is to re-define the DML (\emph{Data Manipulation Language}). e.g., the \emph{update} sentence is redefined as two sentences an update and a create sentence as illustrated in the following example.

@@

\subsection{\label{Comm-temp}Commercial Temporal Database Systems}
Several commercial temporal DBMS exist. Table~\ref{table:commercial-temporal-db} gives an overview of some of the more well-known temporal DBMS and provides references for more information.

Oracle workspace manager~\cite{oracle2009} and TimeDB~\cite{timedb2005} are libraries for dealing with time in OracleDB. On another note, TimeDB and Postgree Temporal~\cite{posgree2009} are similar: both are simple implementations that implement a subset of the Allen operators and some operations for the creation and manipulation of temporal attributes (valid-time, transaction-time or both times are supported). Teradata~\cite{teradata2011} is mainly a business intelligence system designed for data mining. Secondo~\cite{Dieker2000} is an extensible database system in which the core of the database may be replaced by a customized algebra. It is designed for non-standard applications and it supports both valid and transaction-times.

@@

%\subsubsection{Oracle Workspace Manager}
%
%\begin{figure}
%\centering
%\includegraphics[scale=0.5]{graphs/workspaceTopology.eps}
%\caption{A sample topology in the workspace manager.}
%\label{fig:workspace-topology}
%\end{figure}
%
%Oracle workspace manager \cite{OraE118602} package allows to get several versions of the data in the same database. It is also possible to version only a table. The main benefits of this are two:
%
%\begin{itemize}
%\item
%Organization and optimization of data in a hierarchical frames: It is possible to create a frame of time (a workspace) and work only with that partial version of the data. It is provided a mechanism to merge data among different workspaces and to solve conflicts. The organization of data in these workspaces for very large tables result in smaller time access.
%\item
%Valid time as well as transaction-time are managed by the system. Each update sentence makes a change in the versioning of a row.
%\end{itemize}
%
%When a table is versioned, the system creates a few tables and views as well as auxiliary structures to allow keep several version of the data, while keeping the primary key defined by the user.
%
%A workspace is a logical group of a set of changes (versions of tuples) and allows consistent access, thus the user always obtain the correct data version. Workspaces are ordered in a hierarchy. The top level is called LIVE workspace (see figure \ref{fig:workspace-topology}).  \\
%
%The system only makes a copy of a tuple if it is changed. In order to access different versions of the tuple, the context must be changed. Also, it is not necessarily to modify the SQL code to access to versioned tables.\\
%%When a table is versioned, it is renamed as table\_LT. Here is stored all the data for the information together metadata for the versioning. An auxiliary table is created with the workspace metadata with the name table\_AUX. A view is created with the name of the original table to allow querying without versioning.
